---
title: Erdogan's Popularity - I
author: Sabri Ciftci
date: '2021-04-27'
slug: erdogan-s-popularity-i
categories:
  - Turkey
  - Foreign Policy
  - Erdogan
tags:
  - Turkish Foreign Policy
subtitle: ''
summary: ''
authors: []
lastmod: '2021-04-27T09:24:47-05:00'
featured: no
image:
  caption: 'Image Credit: [**Presidency of Republic of Turkey**](https://www.tccb.gov.tr/assets/resim/album/ozel/cumhurbaskanligi-album-025.jpg)'
  focal_point: ''
  placement: 2
  preview_only: false
projects: []
---



In a recent Arab Pulse [**post**](https://www.arabbarometer.org/2021/04/whither-erdogan-regional-leadership-and-the-struggle-for-arab-hearts-and-minds/), Abdul-Wahab Kayyali has pointed to Erdogan's popularity in the Arab World relative to Mohammad Bin Salman and Ali Khamenei. Kayyali's post relies on data from the sixth wave of the  👉 [**Arab Barometer**](https://www.arabbarometer.org), which is being fielded as of Spring 2021. He makes observations about Erdogan's popularity concerning his electoral legitimacy, his somehow credible leadership claim in the Muslim world, and increased economic as well as cultural interactions between Turkey and the Arab World. This is a timely subject that needs further elaboration, one that I will undertake in several posts.

Foreign policy is a complex business. In addition to domestic political calculations, ideology, religion, international strategy, public opinion, and a nation's capabilities and ambitions all play a role in its formulation. Erdogan's popularity in the Middle East and arguably in the broader Muslim world should be understood within this complex play. When it comes to Erdogan and the Middle East, we need to consider a multitude of factors to make sense of Turkish foreign policy. Erdogan wants to make Turkey an economic powerhouse and a global actor. He has deployed religion, culture, tourism, humanitarian aid, and trade among other factors to reach these ambitions. This strategy seems to have paid off. The balance sheet is on the positive side for Erdogan and Turkey's international image among citizens of the Middle East. Turkey now enjoys a degree of soft power influence and a positive international image in the Middle East (Though, one should note that Turkey's use of hard power strategies is on the rise as Turkey becomes entrenched in various conflicts in the Middle East and struggles in the Eastern Mediterranean). Data from the fifth wave of the [**Arab Barometer**](https://www.arabbarometer.org) shows that Erdogan managed to gain approval for his foreign policy. Figure 1, however, shows that this approval is not universal (I need to perfect this figure, but the wheel will do it for now). When asked about Turkey's foreign policy under Erdogan's leadership, citizens in Libya, Egypt, and Iraq appear to be less enthusiastic. A strong endorsement is observed in Jordan, Palestine, Yemen, and Morocco. 



```{r dataprep, include=FALSE}
library(tidyverse)
library(foreign)
library(expss)
ab5<-as_tibble(read.dta("/Users/sabri/Dropbox/ABprojects/tr.image/ABV_Release_Data.dta"))
ab5s<-select(ab5, country, Q725_1, Q725_2, Q725_3, Q505A, Q1016, Q706B)
ab5s=apply_labels(ab5s, Q725_3="Erdogan's Foreign Policy", country="Country",
                  Q706B="Anti-Americanism", Q505A="Religious Party Support")
ab5s<- ab5s %>% dplyr::na_if("don't know")
ab5s<- ab5s %>% dplyr::na_if("refused")
ab5s
```

```{r, echo=FALSE}
f1<-ab5s %>% drop_na(Q725_3) %>% ggplot(aes(country, fill=Q725_3)) + geom_bar() + labs(title="Figure 1. Evaluation of Erdogan's foreign policy") +
    coord_polar(start=-100) + theme(legend.title = element_text(colour="blue", size=10,  face="bold")) + 
    guides(fill=guide_legend(title="Erdogan's Foreign Policy"))
f1
```


Neo-Ottomanism or Turkish soap operas are frequently mentioned as factors leading to Erdogan's popularity among Arab citizens. However, I want to believe that citizens in Arab-Majority countries are much more sophisticated than simply being responsive to soap operas. While neo-Ottomanism, humanitarian aid, and Turkish soap operas have been significant factors, their utility is far more pronounced for Erdogan's domestic audience then the international audience (I will say probably, because we lack good data to empirically validate this claim). Besides, these elements were the subject of ideological divisions among Turkish citizens, and one can imagine not all Arab citizens will be enthusiastic about the grandeur of the Ottoman Empire or the culture depicted in soap operas which is perceived to be corrupt and hyper-Westernized for the taste of traditionally-minded people. Research has cited structural factors including trade relations, secular institutions, modernization, military might, and the attractiveness of EU membership as possible explanatory factors. In addition, in a co-authored paper 👉 [**Here**](https://doi.org/10.1111/fpa.12090), I argued that other factors including religious worldviews and anti-Americanism may be driving favorable views of Turkey ([**Here is a summary of findings in Turkish**])(https://www.orsam.org.tr/tr/islam-mezhep-ve-anti-amerikancilik-kiskacinda-ortadogu-da-dis-politika-mucadelesi/). The essence of the argument is this: In addition to how citizens perceive Turkey (secular, modern, Western, somehow democratic, etc.), their preferences stemming from their own social paradigm likely shape their views of Erdogan and Turkey. Two such preferences are religious outlooks and opinions of  international actors. This is as complex as it gets, but it is the nature of Turkish foreign policy that is formed at the intersections of religion, history, international and regional factors, soft power strategies, and domestic political calculations.

Luckily, the fifth wave of the [**Arab Barometer**](https://www.arabbarometer.org) includes some questions inquiring about anti-Americanism and religious worldviews, allowing us to tackle these dimensions vis-a-vis the favorability of Turkish foreign policy. Islam is a formidable social force in the Middle East and it is likely that it informs political views. Because Erdogan appeals to Islamic identity and his party presents a model accommodating religion in politics, it is likely that religious outlooks will influence orientations toward Turkey's foreign policy.  While the Arab Barometer includes many questions about personal piety and religious worldviews, one question about religious party support presumably captures the role of personal piety in shaping political outlooks (again, wave 5). Respondents were given two statements asking whether they prefer (i) a religious party over a non-religious party and (ii) a non-religious party over a religious party.  Here is how the responses to this question are distributed along the preferences for Erdogan's foreign policy: 

```{r, echo=FALSE}
ab5s$religiousparty<-dplyr::recode(ab5s$Q505A, "strongly agree (S1)"="Support Religious Party", "agree (S1)"="Support Religious Party",  
                                   "agree (S2)"="Don't Support Religious Party",  "strongly agree (S2)"= "Don't Support Religious Party")

f3<-ab5s %>% drop_na(c("religiousparty", "Q725_3")) %>% ggplot(aes(Q725_3)) + geom_bar(aes(fill=(religiousparty), y = (..count..)/sum(..count..)),  position="dodge")  + labs(title="Figure 2: Religious Party Support and Erdogan's Foreign Policy", x="Erdogan's Foreign Policy", y="Percent") + scale_fill_discrete(name= "Religious Party", labels = c("Support Religious Party", "Don't Support Religious Party")) + coord_flip()

f3
```

Clearly, religious party supporters are more likely to view Turkey's foreign policy favorably relative to non-religious party supporters. Furthermore, those who do not support religious parties tend to view Erdogan-led foreign policy in negative terms. Religion is a potent force shaping attitudes and behavior. In this case, it exerts its influence via religious party support and perceptions of international actors. In fact, in the immediate aftermath of the Arab Spring and following the Egyptian military coup in 2013, Erdogan appealed to the supporters of Islamist parties to present his own party as a model. In my research, I have also obtained a figure that shows the distribution of religious party support and evaluation of Turkish foreign policy across countries, but I am still skeptical about this figure. Some cases make sense, while others seem counter-intuitive. Thus, I would like to see some careful research tackling cross-country variation in religious party support and views about international actors before reaching a conclusion. 

Finally, anti-Americanism has some sway over perceptions of international actors. A leader who can stand against Western powers, specifically to the US, is set to obtain popular support in the Middle East, both domestically and internationally. In the past, leaders like Gamal Abdel Nasser, Khomeini, and Saddam Hussein exploited anti-Western and anti-American sentiments to win the hearts and minds of ordinary citizens. My own research [**( 👉 here)**](https://doi.org/10.1111/fpa.12090) finds that anti-Americanism helps create a positive image for countries like Iran. Erdogan has used this card as well, especially against Israel. He publicly took on Israel in public meetings, championed the Palestinian cause, and criticized the US for its double standard in Israeli-Palestinian conflict to boost his image as the leader of the Muslim world. This strategy worked to provide domestic electoral dividends for his party and positive political capital in the Arab Street for Turkey. The fifth wave of the Arab Barometer includes a question about whether violence against the US may be justified because of their involvement in the region or not. Figure 3 reports the summary of responses to this question and provides some insights about anti-Americanism dimension.

```{r, echo=FALSE}
f2<-ab5s %>% drop_na(c("Q706B", "Q725_3")) %>% ggplot(aes(Q725_3)) + geom_bar(aes(fill=(Q706B), y = (..count..)/sum(..count..)), 
 position="dodge")  + labs(title="Figure 3: Anti-Americanism and Turkish Foreign Policy", x="Erdogan's Foreign Policy", y="Percent")+ theme(axis.text.x = element_text(angle = 90)) + scale_fill_discrete(name= " ",    labels = c("Violence against the US may be justified because of their actions", 
    "Violence against the US is never justified", "Neither Statement")) + theme(legend.position = "bottom") + facet_wrap(~country)
f2
```

The data shows a correlation between favorable ratings of Turkish foreign policy and justification of violence against the US targets in countries with the most favorable ratings of Turkey's foreign policy. This correlation is especially visible in Jordan, Palestine, and Yemen, where positive ratings of foreign policy overlap with higher levels of anti-Americanism. In countries where thepublic is less positively inclined toward Erdogan's foreign policy, we see the opposite pattern. 

Overall, the positive international image of Turkey in the Middle East stems from two sources. First, _what Turkey is_ may inform perceptions about its foreign policy. Turkey's secular, modern, and somewhat democratic record along with its soft power strategies including humanitarian policies likely boost its international image. Second, domestic cleavages along religious outlooks and opinion of international actors (i.e. anti-Americanism) may shape public opinion toward Erdogan and Turkey. Erdogan has exploited both fronts to make Turkey a role model and carve a leadership role for himself in the Arab world. 
